---
title: "oSCC Analysis"
author: "Elijah Willie"
date: "2023-06-27"
output: html_document
---

```{r}
rm(list = ls())
suppressPackageStartupMessages({
  library(FuseSOM)
  library(spicyR)
  library(lisaClust)
  library(scater)
  library(ClassifyR)
  library(scFeatures)
  library(lme4)
  library(lmerTest)
  library(cytomapper)
  library(SKM)
  library(cvAUC)
  library(dplyr)
  library(tidyverse)
  library(ggsci)
  library(caret)
  library(e1071)
  source('~/Documents/PhD/oSCC_Analysis/scripts/CustomCV.R')
})
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = T) 
# knitr::opts_chunk$set(fig.width=10, fig.height=6) 
```

```{r}
nCores <- 4
BPPARAM <- simpleSeg:::generateBPParam(nCores)
theme_set(theme_classic())
```

# load the normalized data, images and metadata
```{r}
load("../Data/oSCC_cells_normalised.RData")
```

## Perform the clustering using FuseSOM
```{r}
set.seed(1994)
markers <- markers[!markers%in%c("HH3", "DNA1", "DNA2")]

useMarkers <- c("CD20", #B cells
                "CD31",  #Endothelial cells
                "podoplanin", #Epothelial cells
                "CD66a", # Granulocytes
                "FX111A",  # Macrophages
                "CD14", # Monocytes
                "CD3", "CD4", "CD8a", # T cells
                "CD56", # NK cell
                "CD11c", # Dendritic cells
                # "MPO", # Neutrophils
                "CD45", # Immune
                "panCK","p40", # Tumour
                'FoxP3')
                
secMarkers <- c('CD68', 'HLADR', 'MPO', 'CD103', 'CD16', 'Tbet', 'CD197', 
                'CD45RA', 'CD45RO', 'DC-SIGN', 'CADM1', 'PD1')
# Generate SOM and cluster cells into 20 groups.
cells <- runFuseSOM(cells,
                    markers = useMarkers,
                    assay = 'norm',
                    numClusters = 13)
```

# Check how many clusters should be used.
We can check to see how reasonable our choice of 20 clusters is using the `estimateNumCluster` and the `optiPlot` functions. Here we examine the Gap method, others such as Silhouette and Within Cluster Distance are also available.
```{r, eval=FALSE}
# Generate metrics for estimating the number of clusters.
# As I've already run runFuseSOM I don't need to run generateSOM().
cells <- estimateNumCluster(cells, kSeq = 2:30)

# 'jump', 'slope', 'wcd', 'gap', 'silhouette'
optiPlot(cells, method = "silhouette")
```

# Generat heatmaps of the initial clustering
```{r}
# Visualise marker expression in each cluster.
scater::plotGroupedHeatmap(cells, 
                           features = useMarkers, 
                           group = "clusters", 
                           block = "clusters",
                           exprs_values = "norm",
                           center = TRUE, 
                           scale = TRUE, 
                           zlim = c(-3,3),
                           cluster_rows = FALSE)
```
## Annotate the clustering based on marker expression
```{r}
colData(cells)$clusters_13 <- dplyr::recode(colData(cells)$clusters,
                                         'cluster_1' = 'Squamous Cell 1',
                                         'cluster_2' = 'Squamous Cell 2',
                                         'cluster_3' = 'CD8+ T cell',
                                         'cluster_4' = 'B cell',
                                         'cluster_5' = 'CD4+ Treg',
                                         'cluster_6' = 'Squamous Cell 3',
                                         'cluster_7' = 'Endothelial cell',
                                         'cluster_8' = 'Granulocyte',
                                         'cluster_9' = 'Neutrophil',
                                         'cluster_10' = 'Mo-Derived DC',
                                         'cluster_11' = 'Squamous Cell 4',
                                         'cluster_12' = 'Fibroblast',
                                         'cluster_13' = 'Stromal Fibroblast')
```

## Visualize annotation
```{r}
# pdf(file = "Plots/key_markers.pdf",   # The directory you want to save the file in
#     width = 8, # The width of the plot in inches
#     height = 6) # The height of the plot in inches

# Visualise marker expression in each cluster.
scater::plotGroupedHeatmap(cells, 
                           features = useMarkers, 
                           group = "clusters_13", 
                           block = "clusters_13",
                           exprs_values = "norm",
                           center = TRUE, 
                           scale = TRUE, 
                           zlim = c(-3,3),
                           cluster_rows = FALSE)
# dev.off()
```
## Visualize all markers
```{r,fig.height=6, fig.width=8}
# pdf(file = "Plots/All_markers.pdf",   # The directory you want to save the file in
#     width = 8, # The width of the plot in inches
#     height = 10) # The height of the plot in inches

# Visualise marker expression in each cluster.
scater::plotGroupedHeatmap(cells, 
                           features = markers, 
                           group = "clusters_13", 
                           block = "clusters_13",
                           exprs_values = "norm",
                           center = TRUE, 
                           scale = TRUE, 
                           zlim = c(-3,3),
                           cluster_rows = FALSE)
# dev.off()
```

# Test For association between the proportion of each cell type and progression status
The `colTest` function allows us to quickly test for associations between the proportions of the cell types and progression status using either wilcoxon rank sum tests or t-tests. Here we see a p-value less than 0.05 but this does not equate to a small fdr.

```{r}
testProp <- colTest(cells, 
                      feature = "clusters_13",
                      condition = "Responder", type = 'wilcox')
testProp 
```

# Perform dimension reduction using UMAP
```{r, fig.height=6, fig.width=8}
set.seed(51773)
cells <- scater::runUMAP(cells,
                         subset_row = useMarkers,
                         exprs_values = "norm",
                         BPPARAM = BPPARAM, 
                         n_threads = 7, 
                         n_neighbors = 15,
                         metric = 'cosine',
                         verbose = TRUE)

df_umap <- reducedDim(cells, 'UMAP') %>%
    as.data.frame()
colnames(df_umap) <- c('UMAP 1', 'UMAP 2')

df_umap$`Cell Type` <- cells$clusters_13


# pdf(file = "Plots/Umap_plot.pdf",   # The directory you want to save the file in
#     width = 8, # The width of the plot in inches
#     height = 6) # The height of the plot in inches

df_umap %>% ggplot(aes(`UMAP 1`, `UMAP 2`)) +
    geom_point(aes(color = `Cell Type`), size = 2) + 
    ggtitle('Umap plot of cells') +
  theme_bw() +
    theme(axis.text.x = element_text(size = 12),
        plot.title = element_text(size = 15, hjust = 0.5),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)
        ) + scale_color_d3(palette = 'category20') +
  guides(colour = guide_legend(override.aes = list(size=6)))

# dev.off()
```

# use SpicyR to test for changes in pairwise spatial relationships between cell types.
```{r}
spicyTest <- spicy(cells, 
                   condition = "Responder", 
                   cellType = "clusters_13",
                   imageID = "imageID",
                   subject = "sampleID",
                   spatialCoords = c("m.cx", "m.cy"),
                   Rs = c(20, 50, 100),
                   sigma = 50,
                   BPPARAM = BPPARAM,
                   weights = TRUE)
topPairs(spicyTest, n = 10) 
```

## Generate the SpicyR plot
```{r,fig.height=4, fig.width=6}
# pdf(file = "Plots/SpicyR_plot.pdf",   # The directory you want to save the file in
#     width = 8, # The width of the plot in inches
#     height = 6) # The height of the plot in inches
# # Visualise which relationships are changing the most.
p <- signifPlot(spicyTest,
           breaks = c(-1.5, 2, 0.5))
p$layers[[6]]$data$condition <- c('NR', 'R')
shape.legend = c(NR = "\u25D6", R = "\u25D7")
p + scale_shape_manual(values = shape.legend)

# dev.off()
```

# Use lisaClust to Cluster cells into spatial regions with similar composition.
```{r}
set.seed(51773)
cells <- lisaClust(cells, 
                   k = 5, 
                   Rs = c(20, 50, 100),
                   sigma = 50,
                   spatialCoords = c("m.cx", "m.cy"), 
                   cellType = "clusters_13",
                   BPPARAM = BPPARAM)
```

## Visualize a sample image with the regions
```{r, fig.height=6, fig.width=8}
# Extract cell information and filter to specific image.
df <- colData(cells) |>
as.data.frame() |>
dplyr::filter(imageID == unique(imageID)[4])
# Colour cells by their region.
ggplot(df, aes(x = m.cx, y = m.cy, colour = region)) +
  geom_point() + ggtitle('Localisation across a single image') + 
  theme(plot.title = element_text(hjust = 0.5))

```

## Use hatching to visualise regions and cell types.
```{r, fig.height=6, fig.width=8}
# pdf(file = "Plots/Hatching_plot.pdf",   # The directory you want to save the file in
#     width = 8, # The width of the plot in inches
#     height = 6) # The height of the plot in inches
hatchingPlot(cells,
             useImages = unique(cells$imageID)[4],
             cellType = "clusters_13",
             spatialCoords = c("m.cx", "m.cy"),
             window = "concave",
             nbp = 300,
             line.spacing = 41) + 
  
  scale_region_manual(values = c(region_1 = 1,
                               region_2 = 2,
                               region_3 = 3,
                               region_4 = 4,
                               region_5 = 5)) + 
  
   guides(colour = guide_legend(ncol = 2))

# dev.off()
```
## Test for association with responding
```{r}
testRegion <- colTest(cells, 
                      feature = "region",
                      condition = "Responder",
                      type = 'ttest')
testRegion %>%
  knitr::kable()
```

## Plot boxplots of responder proprotions in each region
```{r,fig.height=5, fig.width=8}
x <- unique(cells$imageID)
cellPropRegion <- getProp(cells,
                    feature = "region", imageID = "imageID", logit = F)

cellPropRegion$Responder <- clinicalData[match(x,clinicalData$sampleID), "Responder"]

cellPropRegionMelted <- data.table::melt(cellPropRegion)

colnames(cellPropRegionMelted) <- c('Responder', 'Region', 'Proportion')

cellPropRegionMelted <- cellPropRegionMelted %>%
    na.omit()


cellPropRegionMelted$Region <- dplyr::recode(cellPropRegionMelted$Region,
                                            'region_1' = 'Region 1',
                                            'region_2' = 'Region 2',
                                            'region_3' = 'Region 3',
                                            'region_4' = 'Region 4',
                                            'region_5' = 'Region 5')

cellPropRegionMelted$Responder <- dplyr::recode(cellPropRegionMelted$Responder,
                                                'Y' = 'R',
                                                'N' = 'NR'
)

# pdf(file = "Plots/Region_proportion_plot.pdf",   # The directory you want to save the file in
#     width = 8, # The width of the plot in inches
#     height = 5) # The height of the plot in inches

cellPropRegionMelted %>% ggplot(aes(Responder, Proportion)) +
    geom_boxplot(aes(fill = Responder)) + 
    geom_jitter() +
    facet_grid(~Region) +
    ggtitle('Boxplot of region proportions for responder and non-responder') +
    theme(axis.text.x = element_text(size = 15),
        plot.title = element_text(size = 12, hjust = 0.5),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        legend.text = element_text(size = 12),
        strip.text.x = element_text(size = 12),
        legend.title = element_text(size = 12)
        ) + scale_fill_jama()

# dev.off()
```
## Region - cell type enrichment heatmap
```{r, fig.height=4, fig.width=6}
# pdf(file = "Plots/Region_map_plot.pdf",   # The directory you want to save the file in
#     width = 6, # The width of the plot in inches
#     height = 4) # The height of the plot in inches

# Visualise the enrichment of each cell type in each region
regionMap(cells, cellType = "clusters_13", limit = c(0.2, 5))
# dev.off()
```
# Test for difference between cell means for cell type and region
```{r}
responderDf <- colData(cells) %>%
  as.data.frame()

markerDf = 
  responderDf |> 
  select(imageID, sampleID, clusters_13, region) |> 
  cbind(data.frame(t(assay(cells, "norm")))) |>
  select(imageID, sampleID, clusters_13, region, useMarkers)

# Mean marker per cell type
markerMeanCellType = markerDf |> 
  select(-c(sampleID, region)) |> 
  group_by(imageID, clusters_13) |> 
  summarise_at(vars(-group_cols()), mean, na.rm = TRUE) |> 
  pivot_longer(-c(imageID, clusters_13), names_to = "markers") |> 
  pivot_wider(names_from = c(clusters_13, markers), values_from = value) |> 
  column_to_rownames("imageID") %>% 
  replace(is.na(.), 0)


# Mean marker per region
markerMeanRegion = markerDf |> 
  select(-c(sampleID, clusters_13)) |> 
  group_by(imageID, region) |> 
  summarise_at(vars(-group_cols()), mean, na.rm = TRUE) |> 
  pivot_longer(-c(imageID, region), names_to = "markers") |> 
  pivot_wider(names_from = c(region, markers), values_from = value) |> 
  column_to_rownames("imageID") %>% 
  replace(is.na(.), 0)
  

# Mean marker per cell type per region
markerMeanCellTypeRegion = markerDf |> 
  select(-c(sampleID)) |> 
  group_by(imageID, clusters_13, region) |> 
  summarise_at(vars(-group_cols()), mean, na.rm = TRUE) |> 
  pivot_longer(-c(imageID, clusters_13, region), names_to = "markers") |> 
  pivot_wider(names_from = c(clusters_13, region, markers), values_from = value) |> 
  column_to_rownames("imageID") %>% 
  replace(is.na(.), 0)
```

## Test marker cell type means
```{r}
testMarkerMeanCellType <- colTest(markerMeanCellType,
                condition = clinicalData[rownames(markerMeanCellType), ]$Responder)
testMarkerMeanCellType |> 
    head(10)
```

## Test marker mean per region
```{r}
testmarkerMeanRegion <- colTest(markerMeanRegion,
                condition = clinicalData[rownames(markerMeanRegion), ]$Responder)
testmarkerMeanRegion |> 
    head(10) %>%
  knitr::kable()
```
## Test marker mean per cell type per region
```{r}
testmarkerMeanCellTypeRegion <- colTest(markerMeanCellTypeRegion,
                condition = clinicalData[rownames(markerMeanCellTypeRegion), ]$Responder)
testmarkerMeanCellTypeRegion |> 
    head(10) %>%
  knitr::kable()
```

# Predictive modelling using Support Vector Machines (SVM) on the marker means per cell types
## Set up the data - use all data for training
```{r}
row_names <- rownames(markerMeanCellType)
colnames(markerMeanCellType) <- janitor::make_clean_names(colnames(markerMeanCellType))

data_final <- markerMeanCellType
colnames(data_final) <- janitor::make_clean_names(colnames(data_final))

data_final$response <- factor(clinicalData[match(row_names,clinicalData$sampleID), 
                                           "Responder"])
data_final <- na.omit(data_final)

n <- ncol(data_final)
train_x <- data_final[, 1:n-1] %>%
  as.matrix()
train_y <- data_final$response
```

## Run 5-fold cross validation with 20 repeats (CV AUC = 0.85)
```{r, eval=FALSE}
cv_dat_svm <- KFoldCustom(train_x = train_x, train_y = train_y, nRepeats = 20,
                      method = 'svm')
```
## Plot the CV ROC curve for all features
```{r, fig.height=6, fig.width=9}
# pdf(file = "Plots/SVM_all_features_CV_plot.pdf",   # The directory you want to save the file in
#     width = 8, # The width of the plot in inches
#     height = 6) # The height of the plot in inches

cv_dat <- cv_dat_svm[[2]]
out <- cvAUC(cv_dat$predictions, cv_dat$labels)

plot(out$perf, col = "black", lty = 3, size = 2,
     main = paste0("Key features 5-fold CV with 20 repeats. Average AUC = ",
                                                      round(out$cvAUC, 2)))
plot(out$perf, col ="red", avg = "threshold", add = TRUE)

# dev.off()
```

## Fit SVM model and predict on the training data
```{r}
model_svm <- support_vector_machine(
  as.matrix(train_x),
  factor(train_y),
  kernel = "linear",
  verbose = F,
  seed = 1994
)

predictions <- predict(model_svm, as.matrix(train_x))
```

## Generate ROC curve from predictions (Perfect prediction on training data)
```{r, fig.height=6, fig.width=8}
roc.pred.test <- ROCR::prediction(predictions$probabilities[, 2], 
                                  train_y)
perf <- ROCR::performance(roc.pred.test, "tpr", "fpr")
  
  
plt_dat = data.frame(
    FPR = perf@x.values[[1]],
    TPR = perf@y.values[[1]]
  )
  
auc_perf <- ROCR::performance(roc.pred.test, measure = "auc")
auc <- auc_perf@y.values[[1]]
  
ggplot(plt_dat, aes(x = FPR, y = TPR)) +
    geom_line(colour = "blue") +
    labs(x = perf@x.name, y = perf@y.name) +
    geom_abline(slope = 1, intercept = 0) + theme_bw() + 
    theme(
      plot.title = element_text(color="Black", size=16, face="bold", hjust = 0.5),
      plot.subtitle = element_text(color = "red", size = 16, hjust = 0.5),
      axis.title.x = element_text(color="Black", size=16),
      axis.title.y = element_text(color="Black", size=16),
      axis.text.y = element_text(size = 16),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 16),
      strip.text.x = element_text(size = 16),
      legend.title = element_text(size=16), #change legend title font size
      legend.text = element_text(size=16) #change legend text font size)
    )  + ggtitle(paste("All featuers AUC =", round(auc, digits = 2)))
```

## Now look at a subset of key 13 features for the SVM model
Again, use all the data for training
```{r}
data_final <- markerMeanCellType %>%
  dplyr::select(squamous_cell_3_pan_ck, cd8_t_cell_cd3, cd4_treg_cd20, neutrophil_cd45,
         fibroblast_cd56, neutrophil_cd31, neutrophil_podoplanin, neutrophil_cd66a,
         neutrophil_cd4, cd4_treg_cd56, neutrophil_cd11c, stromal_fibroblast_fx111a,
         cd4_treg_cd3)

data_final$response <- factor(clinicalData[match(x,clinicalData$sampleID), 
                                           "Responder"])
data_final <- na.omit(data_final)

n <- ncol(data_final)
train_x <- data_final[, 1:n-1] %>%
  as.matrix()
train_y <- data_final$response
```

## Run 5-fold cross validation with 20 repeats (CV AUC = 0.83)
```{r, eval=FALSE}
cv_dat_svm <- KFoldCustom(train_x = train_x, train_y = train_y, nRepeats = 20,
                      method = 'svm')
```
## Plot the CV ROC curve for key features
```{r, fig.height=6, fig.width=9}
# pdf(file = "Plots/SVM_key_features_CV_plot.pdf",   # The directory you want to save the file in
#     width = 8, # The width of the plot in inches
#     height = 6) # The height of the plot in inches

cv_dat <- cv_dat_svm[[2]]
out <- cvAUC(cv_dat$predictions, cv_dat$labels)

plot(out$perf, col = "black", lty = 3, size = 2,
     main = paste0("Key features 5-fold CV with 20 repeats. Average AUC = ",
                                                      round(out$cvAUC, 2)))
plot(out$perf, col ="red", avg = "threshold", add = TRUE)

# dev.off()
```

## Fit SVM model and predict on the training data
```{r}
model_svm <- support_vector_machine(
  as.matrix(train_x),
  factor(train_y),
  kernel = "linear",
  verbose = F,
  seed = 1994
)

predictions <- predict(model_svm, as.matrix(train_x))
```

## Generate ROC curve from predictions (Perfect prediction on training data)
```{r, fig.height=6, fig.width=8}
roc.pred.test <- ROCR::prediction(predictions$probabilities[, 2], 
                                  train_y)
perf <- ROCR::performance(roc.pred.test, "tpr", "fpr")
  
  
plt_dat = data.frame(
    FPR = perf@x.values[[1]],
    TPR = perf@y.values[[1]]
  )
  
auc_perf <- ROCR::performance(roc.pred.test, measure = "auc")
auc <- auc_perf@y.values[[1]]
  
ggplot(plt_dat, aes(x = FPR, y = TPR)) +
    geom_line(colour = "blue") +
    labs(x = perf@x.name, y = perf@y.name) +
    geom_abline(slope = 1, intercept = 0) + theme_bw() + 
    theme(
      plot.title = element_text(color="Black", size=16, face="bold", hjust = 0.5),
      plot.subtitle = element_text(color = "red", size = 16, hjust = 0.5),
      axis.title.x = element_text(color="Black", size=16),
      axis.title.y = element_text(color="Black", size=16),
      axis.text.y = element_text(size = 16),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 16),
      strip.text.x = element_text(size = 16),
      legend.title = element_text(size=16), #change legend title font size
      legend.text = element_text(size=16) #change legend text font size)
    )  + ggtitle(paste("Key features AUC =", round(auc, digits = 2)))
```

## Plot the weights from the SVM model
```{r}
sv <- model_svm[["fitted_model"]][["SV"]]
coef <- model_svm[["fitted_model"]][["coefs"]]

w = t(sv) %*% coef %>%
  as.data.frame()
b = -model_svm[["fitted_model"]][["rho"]]

w$Feature <- rownames(w)
colnames(w)[[1]] <- 'Weight'

# pdf(file = "Plots/SVM_features.pdf",   # The directory you want to save the file in
#     width = 8, # The width of the plot in inches
#     height = 6) # The height of the plot in inches
w$Feature <- gsub('fox_p3', 'foxp3', w$Feature)
w$Feature <- gsub('pan_ck', 'panck', w$Feature)
w$Feature <- stri_replace_last(w$Feature, fixed = "_", ": ") 
w$Feature <- gsub('_', ' ', w$Feature)

ggplot(w, aes(x=reorder(Feature, Weight), y=Weight, fill=Weight))+
   geom_bar(stat="identity") +
  scale_fill_gradient2(low="darkblue", high="darkred") +
        ggtitle("Plot of feature weights for SVM model")+
        labs(x = "Feature",
             y = "Weight") +
        theme(axis.title = element_text(size = 15, face = "bold"),
              axis.text = element_text(size = 15),
              plot.title = element_text(size = 16, hjust = 0.5, face = "bold")) + 
  coord_flip()

# dev.off()
```
## Generate scatter plot of top two features colored by responder
```{r, fig.height=6, fig.width=9}
df_sub <- data_final %>% 
    select(cd4_treg_cd3, neutrophil_cd45) %>%
    as.data.frame()

df_sub$Responder <- train_y
# pdf(file = "Plots/Scatterplot_svm_top_2_features.pdf",   # The directory you want to save the file in
#     width = 9, # The width of the plot in inches
#     height = 6) # The height of the plot in inches

df_sub %>% ggplot(aes(cd4_treg_cd3, neutrophil_cd45)) +
    geom_point(aes(color = Responder), size = 3) +
    scale_color_aaas() +
    theme_bw() +
    ylab('Neutrophil CD45') +
    xlab('CD4 Treg CD3') +
    ggtitle('Mean marker abundance for CD4 Treg: CD3 and Neutrophil CD45') +
    theme(axis.text.x = element_text(size = 16),
          plot.title = element_text(size = 18, hjust = 0.5),
          axis.text.y = element_text(size = 16),
          axis.title.y = element_text(size = 16),
          axis.title.x = element_text(size = 16),
          legend.text = element_text(size = 16),
          legend.title = element_text(size = 16)
    )

# dev.off()

```

## Fit an SVM model on the top two features and plot decision boundary
```{r,fig.height=8, fig.width=9}
svmfit <- svm(Responder~., data = df_sub, kernel = "linear", scale = FALSE,
              probability=TRUE)

plot_svm_jk(df_sub,svmfit, ESL_theme = F) + ylab('Neutrophil CD45') +
    xlab('CD4 Treg CD3') +
    ggtitle('SVM plot of mean marker abundance for CD4 Treg: CD3 and Neutrophil CD45') +
    theme(axis.text.x = element_text(size = 16),
          plot.title = element_text(size = 18, hjust = 0.5),
          axis.text.y = element_text(size = 16),
          axis.title.y = element_text(size = 16),
          axis.title.x = element_text(size = 16),
          legend.text = element_text(size = 16),
          legend.title = element_text(size = 16)
    ) + scale_color_aaas()
  
```
